---
title: "Binary Logistic Regression DrPH (Epidemiology)"
author: "Kamarul Imran M"
date: "8 April 2017"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    keep_md: true

---

\newpage

# Introduction

## Modeling Binary Outcome Data 

The suggested steps are:

1. Describe data
2. Explore data - Exploratory Data Analysis
3. Estimate parameters 
4. Interpret parameters
4. Make Inference
5. Calculate the fitted values 
6. Make Prediction
7. Assessing model
8. Remedy model


# Prepare workspace

## Locate files

* Browse your folders. 
* Look for the files. 
* Check the path to the folder containing the files

## Set the folder  

Set our working directory. REMEMBER! your working directory (working folder) is different from my working directory

```{r, echo=FALSE}
# this is my working directory. You have to specify yours
setwd("E:/Epi_Stat_Matters/LectureNotes2015/binary-logistic/binary-logistic-DrPH-2015/BinaryLogisticDrPH-Epid/Binary_Logistic_DrPH_Epid_Practicals")



```

## Read data

* Read our data in the working folder
* Then, save as a csv file in our working directory

```{r}
mydata <- read.csv('logistic_data.csv', sep = ",", header = TRUE)

```

# Preliminary analysis

## Describe your data

```{r eval=FALSE}
# observe data the first 10 observations 
head(mydata,10)
```

More fancy, we can use 'psych::describe' function

```{r}
library(psych)
describe(mydata)
```

Admit and Rank are taken as numerical variable which does not make sense. 
We need to convert them to categorical (factor) variables.

```{r}
str(mydata$admit)
str(mydata$rank)
```

```{r}
mydata$admit2 <- factor(mydata$admit, labels = c('no','yes'))
mydata$rank2 <- factor(mydata$rank, labels = c('first', 'second', 'third', 'fourth'))
head(mydata)
str(mydata)
```

## Explore your data

Use plots like
* Histogram for numerical variables
* and barplot for categorical variables, at least. 

# Model parameters

## Estimate parameters

* we estimate the logit or the log odds using `glm` function. 
* We used **summary** to see the results stored as the glm model for example **mylogit** 
* We used **coefficients** to examine the regression coefficients
 

```{r}
mylogit <- glm(admit2 ~ gre + gpa + rank2,family = 'binomial'(link = logit),data=mydata)
summary(mylogit)

```

If we would like to see the estimated **beta** then we can use this

```{r}
coefficients(mylogit)

```

To obtain the odds ratios and their 95% CI, we need to exponentiate using *exp* the regression coefficients or the *betas* 


```{r}
exp(coefficients(mylogit))
exp(confint(mylogit))

```

## Interpret parameters 

1.  Interpret the *betas* (the log odds) and their 95% CIs
2.  Interpret the *odds ratios* and their 95% CIs

## Inference

Here, we examine 

1.  the p-values (hypothesis testing) and 
2.  the confidence intervals. 

What is the p-values?

The p-value is defined as the probability of obtaining a result equal to or "more extreme" than what was actually observed, when the null hypothesis is true. In frequentist inference, the p-value is widely used in statistical hypothesis testing, specifically in null hypothesis significance testing.
Ref: <https://en.wikipedia.org/wiki/P-value>

For example, suppose that a vaccine study produced a P value of 0.04. This P value indicates that if the vaccine had no effect, you’d obtain the observed difference or more in 4% of studies due to random sampling error.
Reg: <http://blog.minitab.com/blog/adventures-in-statistics-2/how-to-correctly-interpret-p-values>

* First, using the method of maximum likelihood 
* Next, using the SE method (function **confint.default**)

```{r}
confint(mylogit)
confint.default(mylogit)
```

# Fitted and predicted values

## Fitted values

The fitted values are the expected values of the model. These expected values are the predicted probability for each observation (each patient) in the dataset. You can use 2 functions of getting the fitted values

```{r fitted,  results = 'hide'}
fitted(mylogit)
predict(mylogit, type = 'response')
```

Manually, we can do this to calculate (verify) the conditional probability of being admitted for the current dataset

```{r}
head(fitted(mylogit))

# calculate the logistic probability for the 1st observation
exp(-1.567)/(1+exp(-1.567))
```



# Predicted values

In R, you can prediction of the outcome based on a set of a new data

## Create a new data to make prediction

Similarly, one of the important objectives in modelling is to perform prediction based on the model using new data. 

We can perform these predictions: 

1. Predict the log odds for having the outcome
2. Predict the conditional probability for having the outcome 

Let us say we have these data

gre = 380
gpa = 3.61
rank = first,second, third, fourth

First, we create a data frame

```{r newdata}
new_datal <- data.frame( gre = 380, gpa = 3.61, rank2 = c('first','second','third', 'fourth'))
new_datal
```

Now, we predict the log odds for being **admitted** for a population with **gre=30**, **gpa=3.61** and for different **rank**

```{r}
pred.logit<-predict(mylogit,newdata = new_datal, type='link')
pred.logit
```

We can confirm this by calculate this (using first and third rank)

```{r calc}
-3.99 + 0.00226*380 + 0.8041*3.61 + 0
-3.99 + 0.00226*380 + 0.8041*3.61 - 1.34
```

Notice, that similarity between **predict(x, type='response')** and **fitted** 
Remember, we can calculate the conditional probability of having the outcome 


# Compare models

We compare a model with **gre** and without **gre**. This is done using the deviance method

```{r}
mylogit2 <- glm(admit2 ~ gpa + rank2, family = 'binomial'(link = logit),data=mydata)
anova(mylogit, mylogit2, test = 'Chisq')
```

The p-value shows that mylogit and mylogit2 are different. It suggests the importance of **gre** at the level of significance of 5%. So should we keep **gre**? Perhaps yes, if we take the p-value as the requirement to assess for variable significance.  

# Model assessment

## Linearity in logits 

**gre** is tested for linearity in logit. **gre** is linear but it is rescaled to produce less decimals 

The linearity of logits is tested using library *mfp* package

```{r}
library(mfp)
mylogit3 <- mfp(admit2 ~ fp(gre)+ gpa + rank2, family = 'binomial'(link = logit),data=mydata,verbose=T)
```

Now, let us check the estimated parameters based on fractional polynomials

```{r}
summary(mylogit3)
mylogit3$fptable

```

## Overall model fitness

To do these diagnostics, you can use *LogisticDx* package. This package produces the diagnostic measures for a binary regression model based on covariate pattern

This package produces the Goodness-of-fit for binomial regression including the Hosmer-Lemeshow GOF test and the ROC curve. Usually, the number of groups (quantiles) equal 10 to perform the Hosmer-Lemeshow test. At the same time, we plot the ROC curve. Similarly, we can check the *LogisticDx::auc* value

The *LogisticDx::dx* produces the estimates of residual diagnostics such as
1.  standardized residuals
2.  dchi-square
3.  ddeviance

```{r}
library('LogisticDx')
dx_mylogit2 <- dx(mylogit2,byCov=T)
head(dx_mylogit2, 10)
```

Use *LogisticDx::gof* to produce 
1.  Hosmer-Lemeshow GOF test
2.  Osius and Rojek's tests
3.  the auc

```{r roc}
fit.mylogit2 <- gof(mylogit2, g=10, plotROC = T)
fit.mylogit2
```


To obtain the contigency table for the Hosmer-Lemeshow GOF test, we can use

```{r}
fit.mylogit2$ctHL
```

and for the GOF test

```{r}
fit.mylogit2$gof
```

We can also perform model fitness by using *ROCR* package

```{r }
library(ROCR)
pred.prob2 <- predict(mylogit2, type='response')
pred.prob22 <- prediction(pred.prob2, mydata$admit2)
pred.prob22f <- performance(pred.prob22, measure='tpr', x.measure='fpr')
plot(pred.prob22f)
```

Using *ROCR* package, we can also calculate the *AUC*

```{r AUC}
auc2<-performance(pred.prob22, measure='auc')
auc2@y.values[[1]]
```

Another package is *MKmisc* package, to perform the Hosmer-Lemeshow test of GOF

```{r gof}
library(MKmisc)
HLgof.test(fit = fitted(mylogit2), obs = mydata$admit)
```


## Diagnostic statistics

We can start of plotting these

```{r plot11}
plot(dx_mylogit2$P, dx_mylogit2$dChisq)
plot(dx_mylogit2$P, dx_mylogit2$dDev)
```

You may try and plot the covariate pattern numbers (identifiers) by using this

```{r covarpat}
library(epiR)
dat.mf2 <- model.frame(mylogit2)
head(dat.mf2)
cv_mf2 <- epi.cp(dat.mf2[-1])
head(cv_mf2$cov.pattern)
```


The *plot* function Will return many diagnostic plots

```{r}
plot(mylogit2)
```

Using K-fold validation. Will not discuss here.


# Logistic model with interaction of predictors

Let use see how we deal an interaction. First, read data from this text file.

Columns (variables) no 2, and from 5 to 10 need to be converted to categorical (factor) variables. We will use *lapply* function for that. 

```{r}
data.l<-read.table("LOWBWT.txt",header=T)
data.l[,c(2,5:10)]<-lapply(data.l[,c(2,5)],factor)
```

Now, observe the first few data

```{r}
head(data.l)
```

To simulate a binary predictor variable, we now recode LWT to LWD (LWT<110 vs >=110)

```{r}
data.l$LWD <- findInterval(data.l$LWT, 110)
head(data.l$LWD)
```

Let us verify our categorization

```{r}
data.l$LWD <- factor(data.l$LWD, labels = c("less 110",">=110"))
head(data.l$LWD, 10)
head(data.l$LWT, 10)
str(data.l$LWD)
```

Model the relationship between the outcome variable (LOW = 0,1) with predictors of LWD and AGE interacting with each other. You may use # to perform that 

```{r}
mod.lwd.age <- glm(LOW ~ LWD*AGE, family = binomial(link =logit ),data=data.l)
summary(mod.lwd.age)
```

Predict our model (with a two-way interaction between age and LWD using a new set of data. We can create such a dataset with this

```{r}
newdata.2<-data.frame(AGE = c(15,15,20,20),LWD= rep(c("less 110",">=110"), 2))
newdata.2
```

Now let us predict the log odds for the model with the interaction term

```{r}
predict(mod.lwd.age, newdata = newdata.2)

```

Can you prove the predicted log odds manually?

```{r}
-1.1696 + 0 + 0.0526*15 + 0
-1.1696 + 1.944*1 + 0.0526*15 - 0.1322*1*15
-1.1696 + 0 + 0.0526*20 + 0
-1.1696 + 1.944*1 + 0.0526*20 - 0.1322*1*20
```


# Resources

1. <http://www.ats.ucla.edu/stat/r/dae/logit.htm>
2. <https://cran.r-project.org/web/packages/HSAUR/vignettes/Ch_logistic_regression_glm.pdf>


